---
title: "analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(devtools)
library(tidyverse)
library(here)
load_all() #note TB not sure this is enough I need to also source()
source(file = "../R/fit_DFE.R")
```

Loads the data and transform the data into a matrix format which is the input format for the fitting functions.
Also removes the data with NA.
```{r}
data("data_DFE")
dim(data_DFE) # just a check on the dimension

# Tranforms the selection coefficients to a matrix format
tmp_s <- data_DFE %>%
  pivot_wider(-c(mean_sd), names_from = "Environment", values_from = "mean_s") %>%
  unite(name, c("Position", "AA"))
s_noNA <- matrix(unlist(tmp_s[-1]), ncol = length(tmp_s)-1, byrow = F, dimnames = list(unlist(tmp_s[,1]), names(tmp_s)[-1]))

# Transforms the sd of the error to a matrix format
tmp_sd <- data_DFE %>%
  pivot_wider(-c(mean_s), names_from = "Environment", values_from = "mean_sd") %>%
  unite(name, c("Position", "AA"))
sd_noNA <- matrix(unlist(tmp_sd[-1]), ncol = length(tmp_sd)-1, byrow = F, dimnames = list(unlist(tmp_sd[,1]), names(tmp_sd)[-1]))
```

Fit parameters with the classic FGM called "Martin" in the function options.
A maximum number of dimensions must be set for the fit of the FGM. This parameter is important for the prediction of the fitness of the mutations across the environments. If n is too high in one of the environment there is too much dimensions in which the phenotypes of the mutants will be positioned compared to the number of environments available in the data. If we have $x$ environment $(e_{1}, ..., e_{x})$, and we want to predict the fitnesses of some mutations in environment $e_{x}$ from their fitnesses in $(e_{1}, ..., e_{x-1})$, we have $x-1$ point of references to use for trilateration of the phenotype in a euclidian space of $x-1-1=x-2$ dimensions.
```{r}

envs <- colnames(s_noNA) #vector with the names of each environment for which the fit must be performed
nmax = ncol(s_noNA)-2 #maximum number of dimensions in the fit of the FGM. 
model = "Martin" #classic FGM form corresponding to a quadratic funtcion for the fitness.
method = "NM" #method for the fit of the likelihood funtion (NM = Nelder-Mead)
round_n = T #the number of dimension n in the FGM must be a natural number. However the fittig function returns real numbers so this option round the dimensons and rerun the fit with a fixed round dimensions from the firts fit.
bounds = list(n = c(0.9, nmax+0.1)) #bound for the parameters. Here only n is bounded between 1 and nmax.
iterlim = 500 #number of iteration in the fitting algorithm (previously 5000)
fit <- list() #initialization of the list containing the fit results
s <- s_noNA #simplification of the names of the variable for readability
sd <- sd_noNA

ss <- split(s, col(s, as.factor = T))
ssd <- split(sd, col(s, as.factor = T))
temp <- sapply(ss,
               simplify = F,
               USE.NAMES = T,
               function(x) which(x > quantile(x, 0.15)))
ss <- mapply(FUN = function(x, y) x[y],
             ss, temp,
             SIMPLIFY = F,
             USE.NAMES = T
             )

ssd <-  mapply(FUN = function(x, y) x[y],
             ssd, temp,
             SIMPLIFY = F,
             USE.NAMES = T
             )


#fit the parameters and also keep the information of the fit about the convergence, the log-likelihood...
fit$raw_parameters <- mapply(FUN = fit_DFE,
                             s = ss,
                             sd = ssd,
                             MoreArgs = list(model = model, method = method, round_n = round_n, bounds = bounds, iterlim = iterlim, nmax = nmax),
                             SIMPLIFY = F,
                             USE.NAMES = T)

#extract only the estimates for ease of use in the following.
fit$parameters <- sapply(X = fit$raw_parameters,
                         simplify = T,
                         USE.NAMES = T,
                         FUN = function(e) e$estimate)

#add the extra parameters of the "Tenaillon"" version of FGM which are fixed in the "Martin" version at Q=2 and alpha=1/2
fit$parameters <- rbind(fit$parameters,
                        alpha = rep(1/2, ncol(fit$parameters)),
                        Q = rep(2, ncol(fit$parameters)))
# split the parameters matrix in a list by parameter for ease of use in the following
fit$parameters <- split(fit$parameters, row(fit$parameters, as.factor = T))
fit$parameters <- sapply(X = fit$parameters,
                         simplify = F,
                         USE.NAMES = T,
                         FUN = function(p) {names(p) <- colnames(s); p})
```

Note by TB  --> need an explanation here ... no idea what is going on ?

```{r}
fit$parameters
```

```{r}
grid <- seq(min(s), max(s), length = 100)
dfe <- sapply(X = envs,
              simplify = F,
              USE.NAMES = T,
              FUN = function(e) {
                dist_estim = tibble(environment = rep(e, length(grid)),
                                    s = grid,
                                    density = dfgm_martin(grid, fit$parameters$n[e], fit$parameters$lambda[e], fit$parameters$so[e]))
              }) %>%
  bind_rows()

ggplot() +
      geom_density(aes(x = mean_s, color = "1"), data = data_DFE) +
      geom_line(aes(x = s, y = density, color = "2"), data = dfe) +
      theme(legend.position = c(1, 0), legend.justification = c(1.1, 0.15)) +
      xlab("selection coefficients") +
      scale_color_manual(name = "DFE", breaks = c("1","2"), labels = c("data", "Fit FGM on 85% of data"), values = c("#000000","#D55E00")) +
      facet_wrap(~ environment, scales = "free")

```

# Exporting the fitted distributions 
We save the `dfe` and `fit` as an rds objects for future extra analysis / use and visuals
```{r}
saveRDS(object = fit, file = "../output/fitted_dfes_details.rds")
saveRDS(object = dfe, file = "../output/fitted_dfes_marginal_densities.rds")
saveRDS(object = data_DFE, file = "../output/data_DFE.rds")

```

